\plaintitle

### 2025-05-07 docs/style/article-body.tex
some restructure after mid-seminar

True random numbers, then, cannot be produced solely through code. These systems require some input that is neither replicable nor reproducible. One method that can be realistically used is the inherently random movement of lava lamps \cite{lavarand}, which is used as a backup source of randomness for Cloudflare\footnote{\href{https://blog.cloudflare.com/randomness-101-lavarand-in-production/}{Cloudflare.com, accessed 2025-03-10}}. Another proposed solution for this is quantum random number generation (\emph{QRNG}) \cite{QRNG}. By reading quantum fluctuations from any given source, for instance an optical signal, the inherent natural unpredictability of said source can be harnessed in order to produce a random number from a state that is nigh impossible to reproduce accurately.

In this project, we will be writing firmware for a quantum number generator, which generates true randomness by reading shot noise from an optical signal. Further details about how this signal is produced will be introduced in section 2 and 3, and builds on the work of Clason \cite{Clason2023}. This optical signal will be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). In turn, these random bits will be processed via Toeplitz extraction \cite{toeplitz} in order to process these bits into random numbers. Some processing has to be done on the microcontrollers themselves in order to ensure that the data is workable, and Toeplitz extraction is a tried and tested method to accomplish this. These random numbers will then be output from the microcontroller to the host computer via USB. This thesis will aim to answer one key research question: How can sampled vacuum fluctuations be processed efficiently in order to output QRNG?

@@

\textbf{Research area 2 (\emph{RA2})}: How can we ensure that the output of random numbers is not limited by our firmware, but rather only limited by the USB transfer speed of the microcontroller or the ADC?

@@

Should these internal ADCs prove too limiting, we propose utilizing MAX11102AUB\footnote{\href{https://www.farnell.com/datasheets/1913106.pdf}{Technical specification for MAX11102AUB, accessed 2025-03-13}} with an effective sample rate of 2 million samples per second (MSPS). This ADC provides a 12 bit sample size, providing roughly 2.86 MB/s of sampled data per second, derived by the following calculation.

\[
Data Rate = \frac{2,000,000_{MSPS} \times 12}{8 \times 1,048,576} \approx 2.86 MB/s \text{\phantom{123} (1)}
\]

@@

\[
T =
\begin{bmatrix}
1 & 0 & 1 & 1 & 0 & 1 & 0 & 1 \\
0 & 1 & 1 & 0 & 1 & 0 & 1 & 1 \\
1 & 1 & 0 & 1 & 0 & 1 & 1 & 0 \\
1 & 0 & 1 & 1 & 1 & 0 & 1 & 0
\end{bmatrix}
\quad
k =
\begin{bmatrix}
1 \\ 0 \\ 1 \\ 1 \\ 0 \\ 1 \\ 0 \\ 1
\end{bmatrix}
\]

\[
h(k) = T \cdot x \mod 2
\]

\phantom{TEMP LINEBREAK}

We then perform XOR multiplication on the matrices:

\[
h(k) =
\begin{bmatrix}
(1 \oplus 0 \oplus 1 \oplus 1 \oplus 0 \oplus 1 \oplus 0 \oplus 1) \\
(0 \oplus 1 \oplus 1 \oplus 0 \oplus 1 \oplus 0 \oplus 1 \oplus 1) \\
(1 \oplus 1 \oplus 0 \oplus 1 \oplus 0 \oplus 1 \oplus 1 \oplus 0) \\
(1 \oplus 0 \oplus 1 \oplus 1 \oplus 1 \oplus 0 \oplus 1 \oplus 0)
\end{bmatrix}
\]

Which finally simplifies to:

\[
h(k) =
\begin{bmatrix}
1 \\
0 \\
1 \\
1
\end{bmatrix}
\]

This allows us to interpret the result as a random 4-bit integer:

\[
h(k) = (1011)_2 = 11_{10}
\]

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound between 2.86 MS/s (for the ADC) and 480 MB/s (for the USB output) in which Toeplitz extraction needs to be executed. Any speeds over 2.86 MB/s allows us to upgrade the ADC iteratively to continue increasing the output speed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

This naive implementation will then be flashed to our microcontrollers, beginning with Teensy 4.1 as this is the more capable of the microcontrollers used for this experiment. Code to measure the execution speed in microseconds will be implemented and tested before we load the naive implementation on said microcontroller. We expect that several implementations may be too resource intensive or have a memory complexity far greater than our cheaper, less capable microcontroller are able to handle, and as such these may not be able to be tested until a few iterations of optimization has occurred.

Each iteration will consist of incremental improvements to the algorithm. Our initial investigation has shown several avenues for improving the throughput of the algorithm, and each planned iteration will be discussed briefly. Depending on the empirical results of these iterations, there may be a need for further optimization iterations other than those listed.

@@

In order to evaluate the efficacy of each iteration, we will test the code by processing our pre-determined bitstreams, saving the processed bits in a binary file and evaluating this file with the command line utility \texttt{ent}. This utility runs a series of tests on the data to determine its level of entropy the file has -- rather, how information dense the file is. A perfect result (\emph{e.g. entirely random}) has an entropy value of \(8.0\), whereas acceptable entropy values for our tests are greater than \(7.99\), denoting a high level of randomness. Furthermore, \texttt{ent} can verify the arithmetic mean of the file, summing all the bits in the file and dividing it by the length of the file. A good distribution of randomness should have an arithmetic mean of \(~127.5\). To ensure each iteration works as expected, we will compare the \texttt{ent} result from each iteration against the original baseline, ensuring that the resulting processed bitstring is identical between iterations. There are a few more measurements shown by \texttt{ent}, and more details regarding these can be found in the utilities manual page\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}.

As discussed in section 3, the hardware used will impose a clear bound on how quickly our implementation needs to process the bits in order to match the speed of the ADC, as well as the output speed of the USB-port, both in \(MB/s\). In order to verify accuracy, the output of random numbers should be identical for each sample bitstring tested. Certain iterations might, however, use larger parts of the bitstring, and thus produce a slightly variable result.

Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (while subtracting the interrupt overhead) provides adequate measurements of an algorithms execution speed. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (as the process for this may vary greatly between controllers). As we will use fixed-size bitstrings for evaluation, we can then derive the throughput of the algorithm in \(MB/s\) as follows:

\[
Throughput*{MB/s} = \frac{DataSize*{bits}}{Execution Time_{ms}} \times
\frac{1}{8} \times \frac{1000}{10^6} \text{\phantom{12}(2)}
\]

This measurement allows us to place the throughput of our algorithm soundly in the bounds imposed on us by the hardware. During implementation, measurements of time complexity may be discussed with regards to the code -- but in the end, the performance of the code as it runs is what will be evaluated most thoroughly.

\subsubsection{5.2 Limitations}\label{limitations}

@@

Worth noting is that this iteration (\_and all following iterations that generate the same result) have achieved entropy scores from \texttt{ent} that are greater than 7.99 for all tested data while ensuring it cannot be compressed further. This ensures that the algorithm has successfully removed as much of the patterns as can be reasonably expected from the raw binary data while maintaining the random nature of the data.

\centering

\begin{tabular}{|c|c|c|c|c|c|c|} \hline \textbf{Bit size} &
\multicolumn{3}{c|}{\textbf{Teensy (µs)}} & \multicolumn{3}{c|}{\textbf{Pico
(µs)}} \\ \hline & Min & Max & Avg & Min & Max & Avg \\ \hline 64 & 13 & 14 &
13.1564 & 103 & 117 & 106.3914 \\ 512 & 788 & 791 & 788.3139 & 5296 & 5311 &
5302.4979 \\ 1024 & 3124 & 3130 & 3124.0580 & 21108 & 21119 & 21111.2163 \\
\hline \end{tabular} \caption{Minimum, maximum and average processing time for
Teensy and Pico.} \label{tab:teensy_pico_min_max_avg}

\textbackslash end\{table\}

>>>>

True random numbers, then, cannot be produced solely through code. These systems require some input that is neither replicable nor reproducible. One method that can be realistically used is the inherently random movement of lava lamps \cite{lavarand}, which is used as a backup source of randomness for Cloudflare\footnote{\href{https://blog.cloudflare.com/randomness-101-lavarand-in-production/}{Cloudflare.com, accessed 2025-03-10}}. Another proposed solution for this is quantum random number generation (\emph{QRNG}) \cite{QRNG}. By reading quantum fluctuation signals from any given source, for instance an optical signal, the inherent natural unpredictability of said signal can be harnessed in order to produce a random number from a state that is nigh impossible to reproduce accurately.

Further details about how this signal is produced and sampled will be introduced in section 2 and 3, and builds on the work of Clason \cite{Clason2023}. This optical signal needs to be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). Some processing of the data has to be done in order to ensure that the data is workable. As such, these random bits will be processed via Toeplitz extraction \cite{toeplitz} in order to process these bits into random numbers. These random numbers will then be output from the microcontroller to the host computer via USB. This thesis will aim to answer one key research question: How can sampled vacuum fluctuations be processed efficiently in order to output QRNG?

@@

\textbf{Research area 2 (\emph{RA2})}: Can we ensure that the output of random numbers is not limited primarily by firmware, but rather limited by the processing power or the USB transfer speed of the microcontroller, alternatively by the ADC?

@@

Should these internal ADCs prove too limiting, we propose utilizing MAX11102AUB\footnote{\href{https://www.farnell.com/datasheets/1913106.pdf}{Technical specification for MAX11102AUB, accessed 2025-03-13}} with an effective sample rate of 2 million samples per second (MSPS). This ADC provides a 12 bit sample size, providing roughly 24 Mbit/s of sampled data per second, derived by the following calculation.

\[
\text{ADC Throughput}
= \frac{2,000,000_{\mathrm{MSPS}}\times12}{1,000,000}
\approx24\ \mathrm{Mbit/s}\phantom{123}(1)
\]

@@

\begin{verbatim}
Where n = number of input bits,
      m = number of output bits:

Let X be the input array (length n)
Let T be the seed array (length n + m – 1)
Let Y be a new array of length m

For i from 0 to m - 1:
  sum = 0
  For j from 0 to n - 1:
    sum = sum + X[j] * T[i + j]
  Y[i] = sum mod 2

Return Y
\end{verbatim}

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound over 24 Mbit/s (imposed by the ADC) in which Toeplitz extraction needs to be executed. Any speeds over 2.86 MB/s allows us to upgrade the ADC iteratively to continue increasing the output speed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

This naive implementation will then be flashed to our microcontrollers, beginning with Teensy 4.1 as this is the more capable of the microcontrollers used for this experiment. Code to measure the execution speed in microseconds will be implemented and tested before we load the naive implementation on said microcontroller. Each iteration will consist of incremental improvements to the algorithm. Our initial investigation has shown several avenues for improving the throughput of the algorithm, and each planned iteration will be discussed briefly. Depending on the empirical results of these iterations, there may be a need for further optimization iterations other than those listed.

@@

In order to evaluate the efficacy of each iteration, we will test the code by processing our pre-determined bitstreams, saving the processed bits in a binary file and evaluating this file with the command line utility \texttt{ent}\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}, which is a simple way of evaluating the randomness of binary data.

As discussed in section 3, the hardware used will impose a clear bound on how quickly our implementation needs to process the bits in order to match the speed of the ADC and the output speed of the USB-port in \(Mbit/s\). In order to verify accuracy, the output of random numbers should be identical for each sample bitstring tested. Certain iterations might, however, use larger parts of the bitstring, and thus produce a slightly variable result.

Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (while subtracting the interrupt overhead) provides adequate measurements of an algorithms execution speed. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (as the process for this may vary greatly between controllers). As we will use fixed-size bitstrings for evaluation, we can then derive the throughput of the algorithm in \(Mbit/s\) as follows:

\[
\text{Throughput}{\mathrm{Mbit/s}}
= \frac{DataSize{\mathrm{bits}}}{ExecutionTime_{\mathrm{ms}}}
\times 10^{-3}
\phantom{12}(2)
\]

This measurement allows us to place the throughput of our algorithm soundly in the bounds imposed on us by the hardware. Plugging in the \(24 Mbit/s\) bound imposed by the ADC with an arbitrarily chosen 64-bit sample size, we can derive the average execution speed in microseconds:

\[
T_{\mathrm{ms}}
= \frac{64}{24}\times10^{-3}\ \mathrm{ms}
\approx 2.667\times10^{-3}\ \mathrm{ms}
=2.667\ \mu\mathrm{s}.
\]

\subsubsection{5.3 Limitations}\label{limitations}

@@

Worth noting is that this iteration (\emph{and all following iterations that generate the same result}) have achieved entropy scores from \texttt{ent} that are greater than 7.99 for all tested data while ensuring it cannot be compressed further. This ensures that the algorithm has successfully removed as much of the patterns as can be reasonably expected from the raw binary data while maintaining the random nature of the data.

\begin{tabular}{|c|c|c|c|c|c|c|} \hline \textbf{Bit size} &
\multicolumn{3}{c|}{\textbf{Teensy ($\mu s$)}} &
\multicolumn{3}{c|}{\textbf{Pico ($\mu s$)}} \\ \hline & Min & Max & Avg & Min &
Max & Avg \\ \hline 64 & 13 & 14 & 13.1564 & 103 & 117 & 106.3914 \\ 512 & 788 &
791 & 788.3139 & 5296 & 5311 & 5302.4979 \\ 1024 & 3124 & 3130 & 3124.0580 &
21108 & 21119 & 21111.2163 \\ \hline \end{tabular} \caption{Minimum, maximum and
average processing time for Teensy and Pico.}
\label{tab:teensy_pico_min_max_avg}
